The Three-HITs Theorem\footnote{This work was partially supported by COST Action CA15123.}

### 2017-03-06 Paper/Paper.tex
Small fixes in spelling and grammar

Philosophically, one sees an inductive type as a type `which is built step by step'.
One starts with the nonrecursive constructors, and then at each step new terms are added by applying the recursive constructors to the previously built terms.
This is explained by a theorem which says that inductive types are initial algebras for a functor \cite{dybjer1994inductive,dybjer2003induction} and another one about how these algebras are obtained \cite{adamek1974free}.
For higher inductive types one would expect a similar result which explains how they are constructed.
However, since equations are also allowed in the definition of a higher inductive type, one would expect that during the construction identifications are made by gluing paths.

The goal of this paper is to formally justify this idea by showing that the higher inductive types defined in \cite{basoldhigher} can be generated from the interval, homotopy pushout and colimit.
More concretely, we will prove the following theorem.
\begin{theorem}[Three-HITs Theorem]
\label{thm:three-hit}
In Martin-L\"of Type Theory extended with an interval object, homotopy pushouts and colimits, we can define for each higher inductive type from \cite{basoldhigher} a type with the same introduction, elimination and computation rules.
\end{theorem}
Also, this generalizes the results in \cite{doorn2016constructing,kraus2014general,kraus2016constructions} where the result is showed for truncations.
With this result the metatheory of higher inductive types could be simplified significantly, because rather than a general class, one only needs to check metatheoretical properties for three HITs.

@@

\section{Preliminaries}
\label{sec:prelims}
Let us briefly recall the scheme of higher inductive types which we shall use \cite{basoldhigher}.
First of all, we need \emph{polynomial functors}.
\begin{definition}
Let $X$ be a variable.
Then a \emph{polynomial} is given by the following grammar.
\[
F, G ::= B : \Type \ | \ X \ | \ F \times G \ | \ F + G
\]
\end{definition}
A possible extension would be to use arbitrary containers as in \cite{abbott2005containers}, but we shall refrain to do so.
With that extension the given proof requires the axiom of choice which generally does not hold in type theory.
One can prove that polynomials are functors, and that given a type family $Y : T \rightarrow \Type$ and a polynomial $F$, we get a lift $\bar{F} : F \> T \rightarrow \Type$.
We will write $F \> f$ to denote the application of a polynomial $F$ to a map $f$.
Also, given a dependent function $f : \prod (x : T), Y \> x$, one can define $\bar{F} \> f : \prod (x : H \> T), \bar{H} \> Y \> x$.
Precise definitions of these are given in \cite{basoldhigher}.

@@

Using constructor terms we give the following scheme of higher inductive types.
\begin{definition}
\label{def:hit}
A \emph{higher inductive type} is defined according to the following scheme
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H$ :=
| $c$ : $A \> H \rightarrow H$
| $p_i$ : $\prod (x : B_i \> H), t_i = r_i$ $\quad (i = 1, \ldots, m)$
\end{lstlisting}
where $A$ and each $B_i$ are polynomials, and each $t_i$ and $r_i$ are constructor terms over $c$ of type $H$ with $x : B_i \> H$ as variable..
\end{definition}

@@

With all these notions we can give the introduction, elimination and computation rules of higher inductive types.
The introduction rules for $H$ as given in Definition \ref{def:hit} are
\begin{minipage}{0.6\textwidth}
\[
c : A \> H \rightarrow H
\]
\end{minipage}
\begin{minipage}{0.4\textwidth}
\[
p_i : \prod (x : B_i \> H), t_i = r_i.
\]
\end{minipage}
We also have an elimination rule for which we use the lifting of constructor terms.
\begin{equation*}
	\AxiomC{$\vdash Y : H \rightarrow \Type$}
	\noLine
	\UnaryInfC{$\vdash f : \prod (x : A \> H), \bar{A} \> Y \> x \rightarrow Y \> (c \> x)$}
%	\noLine
	\AxiomC{$\vdash q_i : \prod (x : B_i \> H) (h_x : \bar{B}_i \> Y \> x), \hatt_i =_{p_j \, x} \hatr_j$}
	\BinaryInfC{$\vdash H\frec(f, q_1, \ldots, q_n) : \prod (x : H), Y \> x$}
	\DisplayProof
\end{equation*}
Let us abbreviate $H\frec(f, q_1, \ldots, q_n)$ by $H\frec$.
The type $H$ also has computation rules for each point $t : A \> H$
\begin{equation*}
H\frec \> (c \> t) \equiv f \> t \> (\bar{A} \> H\frec \> t),
\end{equation*}
and for each $a : B_i \> H$
\begin{equation*}
\apd \> H\frec \> (p_i \> a) \equiv q_i \> a \> (\bar{B}_i \> H\frec \> a).
\end{equation*}
Note that these equalities are definitional rather than propositional.

@@

Next we define the homotopy pushout.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\hpushout$ $(A, B, C : \Type)$ $(f : A \rightarrow B)$ $(g : A \rightarrow C)$ :=
| $\inl$ : $B \rightarrow$ $\hpushout$ $A$ $B$ $C$ $f$ $g$
| $\inr$ : $C \rightarrow$ $\hpushout$ $A$ $B$ $C$ $f$ $g$
| $\glue$ : $\prod (a : A), \inl (f \> a) = \inr (g \> a)$
\end{lstlisting}
Note the similarities with the construction of the pushout \cite{mac2013categories}.
Lastly, we can define the homotopy colimit as a higher inductive type in much the same way.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\hocolim$ $(F : \mathbb{N} \rightarrow \Type)$ $(f : \prod (n : \mathbb{N}, F \> n \rightarrow F(n+1)))$ :=
| $\inc$ : $\prod (n : \mathbb{N}), F \> n \rightarrow$ $\hocolim$ $F$ $f$
| $\com$ : $\prod (n : \mathbb{N}) (x : F \> n), \inc \> n \> x = \inc \> (n+1) \> (f \> n \> x)$
\end{lstlisting}

@@

For higher inductive types one would like to do a similar construction.
The first difference is that instead of starting with nothing and adding each constructor at every step, we start with the nonrecursive constructors, and add recursive constructors at every step.
Also, since extra equalities might be present in the higher inductive type, we need to make identifications during the construction.
Rather than just adding points at every step, we also need to glue the right paths.

To understand what should be done more precisely, let us consider an example.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\mathbb{N}_2$ :=
| $0$ : $\mathbb{N}_2$
| $S$ : $\mathbb{N}_2 \rightarrow \mathbb{N}_2$
| $p$ : $S(S \> 0) = 0$
\end{lstlisting}
The first approximation just has a constructor $0$, and after that we add a constructor for $S \> 0$ to obtain the second approximation.
In the third approximation, which we call $F' \> 3$, we found inhabitants $0$, $S \> 0$, and $S(S \> 0)$, and now we can make the first identification.
To do so, we take the following homotopy pushout
\[
\xymatrix
{
	\one + \one \ar[d]_{0 + S(S \> 0)} \ar[r]^-{\delta} & I^1 \ar[d]\\
	F' \> 3\ar[r] & F \> 3
}
\]
to obtain the actual third approximation $F \> 3$, and we continue our construction with that one.
Note that to glue during the $n$th step, we need refer to elements from the $(n-2)$th step.
So, in order to do the identification of the $n$th step, one needs to refer back to a previous approximation.

@@

If $n$ is clear from the context, then we shall not write it down.

@@

Now we have enough to define the approximator.
During the construction one also needs to pay attention to coherency.
For example, in $P$ we have terms using $c_{\rec}$ at most $k$ times, so in $H_{\Approx} \> P$ we have duplicates of terms using $c_{\rec}$ less than $k$ times.
There will be several coherency problems during the proof, so the pushout will seem rather intimidating at first.

@@

The types $Q$ and $P$ represent $F(n-2)$ and $F(n-1)$ respectively, and the maps $i_l$ and $i_r$ are the inclusion maps into $F(n-1)$.
The new type is extended with constructors and paths given by $\inl \comp \inn_1$ and $\inr \comp \inn_1$ respectively.
Also, there are coherency conditions added and these are given by the second and third component of the left upper corner.
Their use will become clear in the Section \ref{sec:rules} when we show the introduction rules.
Now we can define the approximating sequence.

\begin{definition}
In the setting as described, we simultaneously define a sequence of approximations $F : \mathbb{N} \rightarrow \Type$ to $H$ and maps $f : \prod(n : \mathbb{N}), F \> n \rightarrow F(n+1)$ as follows
\begin{align*}
F \> 0 &= H_{\nonrec},\\
F \> 1 &= H_{\Approx} \> (F \> 0) \> \zero \> (\zero\frec (F \> 0)) \> (\zero\frec (F \> 0))\\
F(n+2) &= H_{\Approx} \> (F(n+1)) \> (F \> n) \> \inl \> \inr \qedhere
\end{align*}
For the maps maps $f \> n : F \> n \rightarrow F(n + 1)$, note that we always have the following sequence of maps
\[
\xymatrix@1
{
	P \ar[r] & (I^1 \times B_i \> P) + P \ar[r] & H_{\Approx} P.
}
\]
Taking $P$ to be $F \> n$, then we have $F (n+1) = H_{\Approx} (F \> n)$, and thus the composition gives the map $F \> n \rightarrow F(n + 1)$.
\end{definition}

@@

Now we will construct the map $A_{\rec} \> (\hocolim \> F \> f) \rightarrow \hocolim \> F \> f$, and by Lemma \ref{lem:colim_commute_poly} it suffices to make a map $\hocolim \> (A_{\rec} \comp F) \> (A \> f) \rightarrow \hocolim \> F \> f$.
For this we use the recursion rule of $\hocolim$, and we start with the following string of maps
\[
\xymatrix
{
		A_{\rec}(F \> n) \ar[r]^-{c_{\rec}'} 
		& H_{\rec}(F \> n) \ar[rr]^{\term \comp \iota_2} &
		& H_{\Con}(F \> n) \ar[r]^-{\inl}
		& H_{\Approx}(F \> n) = F(n + 1)
}
\]
where the map $H_{\rec}(F \> n) \rightarrow H_{\Con}(F \> n)$ is the inclusion and the map $H_{\Con}(F \> n) \rightarrow H_{\Approx}(F \> n)$ is the pushout map.
Composing this map with $\inc$, gives maps $A_{\rec}(F \> n) \rightarrow \hocolim \> F \> f$ for all $n : \mathbb{N}$.

@@

For this we use the eliminator of the homotopy colimit.
First, we need to make for all $n$ a map $I^1 \times B_i \> (F \> n)  \rightarrow \hocolim \> F \> f$.
By definition we have the following string of maps
\[
\xymatrix
{
	I^1 \times B_i \> (F \> n) \ar[r]^-{\inn_1} & I^1 \times B_i \> (F \> n) + F \> n \ar[r]^-{\inr}& F(n+1) \ar[rr]^-{\inc(n+1)} & & \hocolim \> F \> f.
}
\]
Next we need to show that the following diagram commutes.
\[
\xymatrix
{
	I^1 \times B_i \> (F \> n) \ar[r]^-{\inn_1} \ar[d]_{I^1 \times B_i \> (f \> n)} & I^1 \times B_i \> (F \> n) + F \> n \ar[r]^-{\inr} \ar[d]_{I^1 \times B_i \> (f \> n) + f \> n} & F(n+1) \ar[d]^{f(n+1)}\\
	I^1 \times B_i \> (F(n+1)) \ar[r]_-{\inn_1} & I^1 \times B_i \> (F(n+1)) + F(n+1) \ar[r]_-{\inr}& F(n+2) 
}
\]
Note that the left diagram commutes.
For the right diagram we can unpack the definition of $f(n+1)$.
\[
\xymatrix
{
	I^1 \times B_i \> (F \> n) + F \> n \ar[r]^-{\inr} \ar[d]_{I^1 \times B_i \> (f \> n) + f \> n} & F(n+1) \ar[d]^{f(n+1)} \ar[dl]^{\inn_2}\\
	I^1 \times B_i \> (F(n+1)) + F(n+1) \ar[r]_-{\inr}& F(n+2) 
}
\]
On the first coordinate it commutes by definition and on the second it commutes by how $F(n+2)$ was defined.

>>>>

Philosophically, one sees an inductive type as a type `which is built step by step'.
One starts with the nonrecursive constructors, and then at each step new terms are added by applying the recursive constructors to the previously built terms.
This is explained by a theorem which says that inductive types are initial algebras for a functor \cite{dybjer1994inductive,dybjer2003induction} and another one about how these algebras are obtained \cite{adamek1974free}.
For higher inductive types one would expect a similar result which explains how they are constructed.
However, since equations are also allowed in the specification of a higher inductive type, one would expect that during the construction identifications are made by gluing paths.

The goal of this paper is to formally justify this idea by showing that the higher inductive types defined in \cite{basoldhigher} can be generated from the interval, homotopy pushout and colimit.
More concretely, we will prove the following theorem.
\begin{theorem}[Three-HITs Theorem]
\label{thm:three-hit}
In Martin-L\"of type theory extended with an interval object, homotopy pushouts and colimits, we can interpret each higher inductive type from \cite{basoldhigher}. This means that we can define a type with the same introduction, elimination and computation rules.
\end{theorem}
Also, this generalizes the results in \cite{doorn2016constructing,kraus2014general,kraus2016constructions} where the result is showed for truncations.
With this result the metatheory of higher inductive types could be simplified significantly, because rather than a general class, one only needs to check metatheoretical properties for three HITs.

@@

\section{Preliminaries}
\label{sec:prelims}
Let us briefly recall the scheme of higher inductive types which we shall use \cite{basoldhigher}.
First of all, we need \emph{polynomial functors}.
\begin{definition}
Let $X$ be a variable.
Then a \emph{polynomial} is given by the following grammar.
\[
F, G ::= B \ | \ X \ | \ F \times G \ | \ F + G
\]
where $B : \Type$.
\end{definition}
A possible extension would be to use arbitrary containers as in \cite{abbott2005containers}, but we shall refrain to do so.
With that extension the given proof requires the axiom of choice which generally does not hold in type theory.
One can prove that polynomials are functors, and that given a type family $Y : T \rightarrow \Type$ and a polynomial $F$, we get a lift $\bar{F} : F \> T \rightarrow \Type$.
We will write $F \> f$ to denote the application of a polynomial $F$ to a map $f$.
Also, given a dependent function $f : \prod (x : T), Y \> x$, one can define $\bar{F} \> f : \prod (x : H \> T), \bar{H} \> Y \> x$.
Precise definitions of these are given in \cite{basoldhigher}.

@@

Using constructor terms we give the following scheme of higher inductive types.
\begin{definition}
\label{def:hit}
A \emph{higher inductive type} is defined according to the following scheme
\lstset{language=Coq}
\begin{lstlisting}
Inductive $H$ :=
| $c$ : $A \> H \rightarrow H$
| $p_i$ : $\prod (x : B_i \> H), t_i = r_i$ $\quad (i = 1, \ldots, m)$
\end{lstlisting}
where $A$ and each $B_i$ are polynomials, and each $t_i$ and $r_i$ are constructor terms over $c$ of type $H$ with $x : B_i \> H$ as variable.
\end{definition}

@@

With all these notions we can give the introduction, elimination and computation rules of higher inductive types.
The introduction rules for $H$ as given in Definition \ref{def:hit} are
%\begin{minipage}{0.6\textwidth}
\[
c : A \> H \rightarrow H
\]
%\end{minipage}
%\begin{minipage}{0.4\textwidth}
\[
p_i : \prod (x : B_i \> H), t_i = r_i.
\]
%\end{minipage}
We also have an elimination rule for which we use the lifting of constructor terms.
\begin{equation*}
	\AxiomC{$\vdash Y : H \rightarrow \Type$}
	\noLine
	\UnaryInfC{$\vdash f : \prod (x : A \> H), \bar{A} \> Y \> x \rightarrow Y \> (c \> x)$}
%	\noLine
	\AxiomC{$\vdash q_i : \prod (x : B_i \> H) (h_x : \bar{B}_i \> Y \> x), \hatt_i =_{p_j \, x} \hatr_j$}
	\BinaryInfC{$\vdash H\frec(f, q_1, \ldots, q_n) : \prod (x : H), Y \> x$}
	\DisplayProof
\end{equation*}
Let us abbreviate $H\frec(f, q_1, \ldots, q_n)$ by $H\frec$.
The type $H$ also has computation rules for each point $t : A \> H$
\begin{equation*}
H\frec \> (c \> t) \equiv f \> t \> (\bar{A} \> H\frec \> t),
\end{equation*}
and for each $a : B_i \> H$
\begin{equation*}
\apd \> H\frec \> (p_i \> a) \equiv q_i \> a \> (\bar{B}_i \> H\frec \> a).
\end{equation*}
Note that these equalities are definitional rather than propositional.

@@

Next we define the homotopy pushout.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\hpushout$ $(A, B, C : \Type)$ $(f : A \rightarrow B)$ $(g : A \rightarrow C)$ :=
| $\inl$ : $B \rightarrow$ $\hpushout$ $A$ $B$ $C$ $f$ $g$
| $\inr$ : $C \rightarrow$ $\hpushout$ $A$ $B$ $C$ $f$ $g$
| $\glue$ : $\prod (a : A), \inl (f \> a) = \inr (g \> a)$
\end{lstlisting}
Note the similarities with the definition of the pushout \cite{mac2013categories}.
Lastly, we can define the homotopy colimit as a higher inductive type in much the same way.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\hocolim$ $(F : \mathbb{N} \rightarrow \Type)$ $(f : \prod (n : \mathbb{N}, F \> n \rightarrow F(n+1)))$ :=
| $\inc$ : $\prod (n : \mathbb{N}), F \> n \rightarrow$ $\hocolim$ $F$ $f$
| $\com$ : $\prod (n : \mathbb{N}) (x : F \> n), \inc \> n \> x = \inc \> (n+1) \> (f \> n \> x)$
\end{lstlisting}

@@

For higher inductive types one would like to do a similar construction.
The first difference is that instead of starting with $\zero$, we start with the nonrecursive constructors.
Also, in this construction, the functor $F$ is applied at every step.
Instead we will add terms built from the recursive constructors using previously constructed terms as arguments.
Since extra equalities might be present in the higher inductive type, we need to make identifications during the construction.
Rather than just adding points at every step, we also need to specified the right paths.

To understand more precisely what should be done, let us consider an example.
\lstset{language=Coq}
\begin{lstlisting}
Inductive $\mathbb{N}_2$ :=
| $0$ : $\mathbb{N}_2$
| $S$ : $\mathbb{N}_2 \rightarrow \mathbb{N}_2$
| $p$ : $S(S \> 0) = 0$
\end{lstlisting}
The first approximation just has a constructor $0$, and after that we add a constructor for $S \> 0$ to obtain the second approximation.
In the third approximation, which we call $F' \> 3$, we found inhabitants $0$, $S \> 0$, and $S(S \> 0)$, and now we can make the first identification.
To do so, we take the following homotopy pushout
\[
\xymatrix
{
	\one + \one \ar[d]_{0 + S(S \> 0)} \ar[r]^-{\delta} & I^1 \ar[d]\\
	F' \> 3\ar[r] & F \> 3
}
\]
to obtain the actual third approximation $F \> 3$, and we continue our construction with that one.
Note that to glue during the $n$th step, we need refer to elements from the $(n-2)$th step.
So, in order to do the identification of the $n$th step, one needs to refer back to a previous approximation.

@@

Note that this type is actually a labeled sum.
This is because $H_{\rec} \> P$ is isomorphic to $A_{\rec} \> P$, and thus this $H_{\Con}^n \> P$ is actually $P + A_{\rec} \> P + \ldots + A_{\rec}^n \> P$.
Hence, we do not need inductive types in general for the construction, but we just need to have sums and products.
If $n$ is clear from the context, then we shall not write it down.

@@

Now we have sufficient to define the approximator.
During the construction one also needs to pay attention to coherency.
For example, in $P$ we have terms using $c_{\rec}$ at most $k$ times, so in $H_{\Approx} \> P$ we have duplicates of terms using $c_{\rec}$ less than $k$ times.
There will be several coherency problems during the proof, so the pushout will seem rather intimidating at first.

@@

The types $Q$ and $P$ represent $F(n-2)$ and $F(n-1)$ respectively, and the maps $i_l$ and $i_r$ are the inclusion maps into $F(n-1)$.
The new type is extended with constructors and paths given by $\inl \comp \inn_1$ and $\inr \comp \inn_1$ respectively.
Also, there are coherency conditions added and these are given by the second and third component of the left upper corner.
Their use will become clear in the Section \ref{sec:rules} when we construct the right introduction rules.
Now we can define the approximating sequence.

\begin{definition}
In the setting as described, we simultaneously define a sequence of approximations $F : \mathbb{N} \rightarrow \Type$ to $H$ and maps $f : \prod(n : \mathbb{N}), F \> n \rightarrow F(n+1)$ as follows
\begin{align*}
F \> 0 &= H_{\nonrec},\\
F \> 1 &= H_{\Approx} \> (F \> 0) \> \zero \> (\zero\frec (F \> 0)) \> (\zero\frec (F \> 0))\\
F(n+2) &= H_{\Approx} \> (F(n+1)) \> (F \> n) \> \inl \> \inr \qedhere
\end{align*}
For the maps maps $f \> n : F \> n \rightarrow F(n + 1)$, note that we always have the following sequence of maps
\[
\xymatrix@1
{
	P \ar[r] & (I^1 \times B_i \> P) + P \ar[r] & H_{\Approx} \> P \> Q \> i_l \> i_r.
}
\]
Taking $P$ to be $F \> n$, then we have $F (n+1) = H_{\Approx} (F \> n)$, and thus the composition gives the map $F \> n \rightarrow F(n + 1)$.
\end{definition}

@@

Now we will construct the map $A_{\rec} \> (\hocolim \> F \> f) \rightarrow \hocolim \> F \> f$, and by Lemma \ref{lem:colim_commute_poly} it suffices to make a map $\hocolim \> (A_{\rec} \comp F) \> (A \> f) \rightarrow \hocolim \> F \> f$.
For this we use the recursion rule of $\hocolim$, and we start with the following string of maps
\[
\xymatrix
{
		A_{\rec}(F \> n) \ar[r]^-{c_{\rec}'} 
		& H_{\rec}(F \> n) \ar[rr]^{\term \comp \iota_2} &
		& H_{\Con}(F \> n) \ar[r]^-{\inl}
		& H_{\Approx}(F \> n) = F(n + 1)
}
\]
Composing this map with $\inc$, gives maps $A_{\rec}(F \> n) \rightarrow \hocolim \> F \> f$ for all $n : \mathbb{N}$.

@@

For this we use the eliminator of the homotopy colimit.
First, we need to make for all $n$ a map $I^1 \times B_i \> (F \> n)  \rightarrow \hocolim \> F \> f$.
By definition we have the following string of maps
\[
\xymatrix
{
	I^1 \times B_i \> (F \> n) \ar[r]^-{\inn_1} & I^1 \times B_i \> (F \> n) + F \> n \ar[r]^-{\inr}& F(n+1) \ar[rr]^-{\inc(n+1)} & & \hocolim \> F \> f.
}
\]
Next we need to show that the following diagram commutes.
\[
\xymatrix
{
	I^1 \times B_i \> (F \> n) \ar[r]^-{\inn_1} \ar[d]_{I^1 \times B_i \> (f \> n)} & I^1 \times B_i \> (F \> n) + F \> n \ar[r]^-{\inr} \ar[d]_{I^1 \times B_i \> (f \> n) + f \> n} & F(n+1) \ar[d]^{f(n+1)}\\
	I^1 \times B_i \> (F(n+1)) \ar[r]_-{\inn_1} & I^1 \times B_i \> (F(n+1)) + F(n+1) \ar[r]_-{\inr}& F(n+2) 
}
\]
Note that the left diagram commutes.
For the right diagram we can unpack the definition of $f(n+1)$.
\[
\xymatrix
{
	I^1 \times B_i \> (F \> n) + F \> n \ar[r]^-{\inr} \ar[d]_{I^1 \times B_i \> (f \> n) + f \> n} & F(n+1) \ar[d]^{f(n+1)} \ar[dl]^{\inn_2}\\
	I^1 \times B_i \> (F(n+1)) + F(n+1) \ar[r]_-{\inr}& F(n+2) 
}
\]
On the second coordinate it commutes by definition and on the first it commutes by how $F(n+2)$ was defined.
